Convergence rate of MAP estimates \\ for the exponential family

### 2020-11-26 main.tex
Add details on mirror descent

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\logpartition}{A}
\newcommand{\conj}{\logpartition^*}
\newcommand{\bregman}{\cB_\logpartition}
\newcommand{\bregmanconj}{\cB_{\logpartition^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\Dir}{\mathrm{Dir}} % Dirichlet
\newcommand{\linear}{\ell} % linearization of a function

@@

\section{Stochastic Proximal Bregman Point}
We see that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp || \natp_0)$. This is a stochastic proximal Bregman step with step-size $\inv{n_0}$. This can also be seen at each step since
\begin{align}
    \hat \natp_{n+1} 
    &= \argmin_\natp -\log p(x_{n+1}|\natp) + (n_0 + n) \bregman(\natp || \hat \natp_n) \\
    &= \argmin_\natp f(\natp) +  \inv{\gamma_n}\bregman(\natp || \hat \natp_n) \; .
\end{align}
Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.

@@

\section{Stochastic Mirror Descent}
It turns out that the MAP can also be seen as the trajectory of stochastic mirror descent
\begin{align}
    \hat \natp_{n+1} 
    &= \argmin_\natp 
    - \langle T(x_{t+1}), \natp \rangle  + \logpartition(\natp) + (n_0 + n) \bregman(\natp || \natp_n) \\
    %&(n_0+n)\left(\logpartition(\natp) - \logpartition(\natp_n) - \langle \nabla \logpartition(\natp_n)  , \natp - \natp_n \rangle \right)  \\
    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpartition(\natp_n) + \langle \nabla \logpartition(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp || \natp_n) \\
    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp || \natp_n)
\end{align}
where $\linear_f(\natp;\natp_n, x_{t+1})$ is the linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$. 
This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpartition$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.

In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpartition$ \citep{bubeck2015convex}.
In our setting this is not always true -- eg gaussians.
However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity. $f$ is $\mu$ strongly-convex and $L$-smooth relative to $h$ if
\begin{align}
	f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y||x)
	\leq f(y) \\
	\text{and} \quad 
	f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y||x), \forall x,y \\
	\iff \mu\cB_h(y||x) \leq \cB_f(y||x) \leq L \cB_h(y||x), \forall x,y \\
	\iff \mu \nabla^2 h(x) \leq \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
\end{align} 
where the last equivalence holds only when $f$ and $h$ are twice differentiable. 
This sweet generalization of smoothness and strong-convexity transfers the Loewner partial order between matrices to functions, via the Hessian. 
As such it can be applied to many functions that were out of reach for $\ell^2$ norm, by taking the appropriate reference function.  For instance $h(x) = -\log(x)$ or $h(x) = x^4$. 
As early as 2011, \citet{birnbaum2011distributed} showed $O(\inv{t})$ convergence rate for mirror descent under smoothness assumption relative to the mirror map. 
More precisely, he proved the simple finite rate
\begin{align}
	f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* || x_0) }{t} \; .
\end{align}
These notions were rediscovered and expanded by \citet{bauschke2017descent} and \citet{lu2018relatively}. If you need to read one, pick \citet{lu2018relatively} -- I found it much much easier and more enjoyable to read.  This latter paper also derived a linear convergence rate for mirror descent under relative smoothness and strong-convexity, with the relative condition number $\frac{L}{\mu}$ appearing.

Now our setting is Stochastic Mirror Descent (SMD), meaning at each step we observe a random unbiased estimate gradient. This setting was studied by \citet{hanzely2018fastest}, who proved in the smooth strongly-convex case with tail averaging :  with constant step-size, linear convergence down to a variance ball, and with step-size $\gamma_t = n_0 + t$ a rate $\tilde O(\inv{t})$. These results match the rates for standard SGD.
This is very interesting for us, but the variance hyper-parameter is oddly defined. Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate. Then the variance bound $\sigma^2$ is a lower bound on the covariance between $g_t$ and $\natp_{t+1}$. 
\begin{align}
	\Cov(g_t,\natp_{t+1}) 
	&= \expect{\langle g_t, \natp_{t+1} \rangle}  - \langle \E[g_t], \E[\natp_{t+1}] \rangle \geq -  \gamma_t \sigma^2, \forall t \\
	 &= \expect{\langle g_t - \nabla f(\natp_t), \natp_{t+1}  - \natp \rangle}, (\forall \natp)
\end{align}
where $\gamma_t$ is the step-size and expectations are conditional on $\natp_t$.
Still need to check €œRelative Continuity€ for Non-Lipschitz Nonsmooth Convex Optimization Using Stochastic (or Deterministic) Mirror Descent, which makes different hypothesis and is well-written.

Also cite "The Information Geometry of Mirror Descent", saying they only give asymptotic results.

\section{Straightforward Convergence Rate}
In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    \expect[x\sim p(.|\natp^*)]{-\log p(x|\natp) + \log p(x|\natp^*)}
    &= \KL( p(.|\natp^*) || p(.|\natp)) \\
    &= \bregman (\natp || \natp^*) \\
    &= \bregmanconj ( \meanp^* || \meanp)
    %& = \logpartition(\natp) - \logpartition(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
\end{align}
where $\conj$ is the entropy, the convex conjugate of the log-partition.
The relationship between Bregman divergences and Fenchel conjugacy is well explained in Wainwright's book, and the article geometry of exponential family. 
The question is: how does this quantity behave when $\natp$ is the MAP estimate ? Can we get bounds -- in expectation or high-probability ?

>>>>

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\logpartition}{A}
\newcommand{\conj}{\logpartition^*}
\newcommand{\bregman}{\cB_\logpartition}
\newcommand{\bregmanconj}{\cB_{\logpartition^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\Dir}{\mathrm{Dir}} % Dirichlet
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size

@@

\section{Stochastic Proximal Bregman Point}
We see that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp || \natp_0)$. This is a stochastic proximal Bregman step with step-size $\inv{n_0}$. This can also be seen at each step since
\begin{align}
    \hat \natp_{n+1} 
    &= \argmin_\natp -\log p(x_{n+1}|\natp) + (n_0 + n) \bregman(\natp || \hat \natp_n) \\
    &= \argmin_\natp f(\natp; x_{n+1}) +  \inv{\gamma_n}\bregman(\natp || \hat \natp_n) \; .
\end{align}
Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.

@@

\section{Stochastic Mirror Descent (SMD)}
\subsection{MAP as SMD}
Let $f(\natp) := \expect[x]{-\log p(x|\natp)} = - \langle \mu_*, \natp\rangle + \logpartition(\natp)$. In words, $f$ is linear modification of a convex function $\logpartition$, which we can access only through noisy estimates of $\mu_*$. It turns out that the MAP estimate can also be seen as the iterates of stochastic mirror descent with mirror map $\logpartition$. First let's recall mirror descent iteration
\begin{align}
	\natp_{t+1} 
	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp || \natp_t)  \\
	& = \nabla \conj (\nabla \logpartition(\natp_t) - \lr \nabla f(\natp_t))
\end{align}
where $\linear_f(\natp; \natp_t) = f(\natp_t) + \langle \nabla f(\natp_t), \natp - \natp_t \rangle$ is the linear approximation of $f$ in $\natp_t$ evaluated at $\natp$. Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpartition(\natp)$, eg computing the convex conjugate of $\logpartition$. 
Note that finding $\natp_*$ is done with 1 step of mirror descent. Indeed plugging in definitions of $f$ and $\mu$ yields
\begin{align}
	&\mu_{t+1}  = \mu_t - \lr (\mu_t - \mu_*) \\
	\implies &\mu_t  = \mu_* + (1- \lr)^t (\mu_0 - \mu_*)
\end{align}
which shows exponential convergence and 1-step convergence when $\lr =1$. Back to our sheep, the MAP iteration can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1})$ a stochastic estimate  of $\nabla f(\natp_t)$
\begin{align}
    \hat \natp_{n+1} 
    &= \argmin_\natp 
    - \langle T(x_{t+1}), \natp \rangle  + \logpartition(\natp) + (n_0 + n) \bregman(\natp || \natp_n) \\
    %&(n_0+n)\left(\logpartition(\natp) - \logpartition(\natp_n) - \langle \nabla \logpartition(\natp_n)  , \natp - \natp_n \rangle \right)  \\
    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpartition(\natp_n) + \langle \nabla \logpartition(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp || \natp_n) \\
    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp || \natp_n)
\end{align}
where $\linear_f(\natp;\natp_n, x_{t+1})$ is the stochastic linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$. This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpartition$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.

\subsection{Relative Smoothness for Mirror Descent}
In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpartition$ \citep{bubeck2015convex}.
In our setting this is not always true -- eg gaussians.
However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity. $f$ is $\mu$ strongly-convex and $L$-smooth relative to $h$ if
\begin{align}
	& f-\mu h \ \text{ convex and } Lh - f \ \text{convex} \\
	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y||x) \leq f(y) \\
	&\text{and} \quad f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y||x), \forall x,y \\
	\iff & \mu\cB_h(y||x) \leq \cB_f(y||x) \leq L \cB_h(y||x), \forall x,y \\
	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
\end{align} 
where the last equivalence holds only when $f$ and $h$ are twice differentiable. 
This sweet generalization of smoothness and strong-convexity transfers the Loewner partial order on symmetric matrices to functions, via the Hessian. 
As such it can be applied to many functions that were out of reach for $\ell^2$ norm, by taking the appropriate reference function.  For instance $h(x) = -\log(x)$ or $h(x) = x^4$. 
As early as 2011, \citet{birnbaum2011distributed} showed $O(\inv{t})$ convergence rate for mirror descent under smoothness assumption relative to the mirror map. 
More precisely, he proved that when $f$ is $L$-smooth relative to $h$, then the suboptimality is upper bounded by the simple formula
\begin{align}
	x_{t+1}  = \argmin_x \langle \nabla f(x_t), x \rangle + \cB_h(x || x_t) \\
	\implies f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* || x_0) }{t} \; .
\end{align}
These notions were rediscovered and expanded by \citet{bauschke2017descent} and \citet{lu2018relatively}. If you need to read one, pick \citet{lu2018relatively} -- I found it much much easier and more enjoyable to read.  This latter paper also derived a linear convergence rate for mirror descent under relative smoothness and strong-convexity, with the relative condition number $\frac{L}{\mu}$ appearing.

\subsubsection{Three point lemma}
These proofs rely on another basic property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 
\begin{align}
	\min	_{x\in C} \overbrace{f(x) + \cB_h(x || y)}^{\phi(x)}
\end{align}
where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$ (we will drop the index, and use a semicolon instead of double bars to lighten the notation), and $y$ is some reference vector. Then 
\begin{align}
	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
\end{align}
This property is an analog of the Pythagorean theorem for generalized projections (setting $f=0$ and $h=\|.\|^2$).

\begin{proof}
	The first order optimality condition is
	\begin{align}
		&\langle \nabla \phi (x_+) , x -  x_+ \rangle \geq 0, \forall x \\
		=& \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  
	\end{align}
	This proof relies on another property called three point property 
	\begin{align}
		\langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  =\cB(x;y) -\cB(x;x_+) - \cB(x_+;y) 
	\end{align}
	which can be proved by expanding the right hand side.
	By convexity of $f$ we also have 
	\begin{align}
		 f(x_+) + \langle \nabla f(x_+), x - x_+ \rangle \leq f(x) \; .
	\end{align} 
	Putting it all together we get
	\begin{align}
		0 
		&\leq  \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
	\end{align}
	which concludes the proof.
\end{proof}

\subsection{Relative smoothness for SMD}
Now our setting is Stochastic Mirror Descent (SMD), meaning at each step we observe a random unbiased estimate gradient. This setting was studied by \citet{hanzely2018fastest}, who proved in the smooth strongly-convex case with tail averaging :  with constant step-size, linear convergence down to a variance ball, and with step-size $\gamma_t = n_0 + t$ a rate $\tilde O(\inv{t})$. These results match the rates for standard SGD.

This is very interesting for us, but the variance hyper-parameter is oddly defined. Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate. Then the variance bound $\sigma^2$ is a lower bound on the covariance between $g_t$ and $\natp_{t+1} = \nabla \logpartition^*(\nabla \logpartition(\natp_t) - \gamma_t g_t )$. 
\begin{align}
	\Cov(g_t,\natp_{t+1}) 
	&= \expect{\langle g_t, \natp_{t+1} \rangle}  - \langle \E[g_t], \E[\natp_{t+1}] \rangle \geq -  \gamma_t \sigma^2, \forall t \\
	 &= \expect{\langle g_t - \nabla f(\natp_t), \natp_{t+1}  - \natp \rangle}, (\forall \natp)
\end{align}
where $\gamma_t$ is the step-size and expectations are conditional on $\natp_t$. In words, $g_t$ and $\natp_{t+1}$ should not be too negatively correlated. In our case, 
\begin{align}
	g_t - \nabla f(\natp_t) = \E[T(X)] - T(x_{t+1})	
\end{align}
so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable.

The other article \citet{lu2019relative} derives another convergence rate for SMD under another gradient variance assumption.

Also \citet{raskutti2015information} only gives asymptotic results.

\section{Straightforward Convergence Rate}
In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    \expect[x\sim p(.|\natp^*)]{-\log p(x|\natp) + \log p(x|\natp^*)}
    &= \KL( p(.|\natp^*) || p(.|\natp)) \\
    &= \bregman (\natp || \natp^*) \\
    &= \bregmanconj ( \meanp^* || \meanp)
    %& = \logpartition(\natp) - \logpartition(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
\end{align}
where $\conj$ is the entropy, the convex conjugate of the log-partition.
The relationship between Bregman divergences and Fenchel conjugacy is well explained in Wainwright's book, and the article geometry of exponential family. 
The question is: how does this quantity behave when $\natp$ is the MAP estimate ? Can we get bounds -- in expectation or high-probability ? 
\begin{align}
	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \\
	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right )
\end{align}
